---
title: "DM-Project-Marechal-Volaris"
format: html
editor: visual
---

# Loading the datasets

```{r}
here::i_am("DM-Project-Marechal-Volaris.Rproj")
library(here)
library(vroom)
counties <- vroom("county_complete.csv")
med_coverage <- vroom("dataset3.csv")
covid <- vroom("coronavirus-covid-19-pandemic-usa-counties.csv")
```
## Cleaning the datasets




# Dataset Presentation.

##The following explanations will concern the content of the three datasets studied in this project

## Data set 1 : "counties"

### Basic informations :

```{r}
# Number of rows and columns
num_rows <- nrow(counties)
num_cols <- ncol(counties)

# Create a table with basic information
data_info <- data.frame(
  Metric = c("Number of Rows", "Number of Columns"),
  Value = c(num_rows, num_cols)
)

# Display the table
print(data_info)
```


### Key variables :

### 1. median_age_2017:Median age of the population in 2017. It reflects the age structure of the population and can influence economic activity and service demand.

### 2. bachelors_2017:Percentage of the population with a bachelor’s degree or higher in 2017. This indicates the educational attainment level, which is closely linked to economic opportunities and income levels.

### 3. hs_grad_2017: Percentage of the population with a high school diploma or equivalent in 2017. It measures basic educational attainment, a critical factor for workforce readiness.

### 4. some_college_2017: Percentage of the population with some college education but no degree in 2017. This reflects the segment of the population with post-secondary education but without completing a formal degree.

### 5. poverty_rate_2017: Percentage of the population living below the poverty line in 2017. It is a key measure of economic hardship and income disparity within the county.

### 6. unemployment_rate_2017: Unemployment rate in 2017. This is a vital indicator of the local labor market’s health and economic stability.

### 7. median_household_income_2017: Median household income in 2017. It provides a measure of the typical income of households in the region, indicating economic well-being.

### 8. female_2017: Percentage of the population identifying as female in 2017. Understanding gender distribution can inform economic and social policy decisions.

### 9. veterans_2017: Number of veterans living in the county in 2017. This highlights the demographic presence of individuals with military service and their potential economic and social needs.

### 10. foreign_born_2017: Percentage of the population born outside the United States in 2017. This variable is critical for understanding immigration patterns and their economic and cultural contributions.

## Data set 2 : "med_coverage"

### Basic informations :

```{r}
# Number of rows and columns
num_observations2 <- nrow(med_coverage)
num_variables2 <- ncol(med_coverage)

# Create a table with basic information
data_info2 <- data.frame(
  Metric = c("Number of Columns (Variables)", "Number of Rows (Observations)"),
  Value = c(num_variables2, num_observations2)
)

# Display the table
print(data_info2)
```


### Key variables :

### 1. Percent of Population with Incomes at or Below 100 Percent of the U.S. Federal Poverty Level Index of Medical Underservice Score: Index related to the percentage of the population below the poverty threshold, used to assess medical needs in the region.

### 2. Percentage of Population Age 65 and Over IMU Score: Score based on the percentage of the population aged 65 and over, reflecting the level of age-related vulnerability.

### 3. Infant Mortality Rate IMU Score: Score associated with the infant mortality rate, used to evaluateh medical needs in underserved areas.

### 4. Percent of Population with Incomes at or Below 100 Percent of the U.S. Federal Poverty Level: Percentage of the population with income equal to or below 100% of the federal poverty level, reflecting poverty levels in the area studied.

### 5. Infant Mortality Rate: Infant mortality rate (deaths per 1,000 live births), a key indicator of overall health and available care in the region.

### 6. Providers per 1000 Population: Number of healthcare providers available per 1,000 residents, measuring access to medical care for the population.

### 7. Percentage of Population Age 65 and Over: Percentage of the population aged 65 and older. This figure helps evaluate the proportion of elderly people in the region, often associated with increased healthcare needs.

### 8. Index of Medical Underservice Score: Score indicating the level of medical underservice. The higher the score, the more underserved the area is considered.

### 9. Primary State Name: Name of the state where the underserved area is located. This helps contextualize designations geographically.

### 10. Common County Name: Name of the county associated with the designated area. This detail is useful for localized or specific analyses.

### 11. MUA/P ID: Unique identifier assigned to each area or professional designated as underserved. It helps differentiate entities in the database.

### 12. MUA/P Area Code: Geographic code representing the underserved area. It is used to link data to specific zones.

### 13. MUA/P Service Area Name: Name assigned to the medically underserved area or service. This provides clear and geographically localized identification.

### 14. Designation Type Code: Indicates the type of designation, such as a medically underserved area. This facilitates classification of designation types.

## Data set 3 : "covid"

### Basic informations :

```{r}
# Number of rows and columns
num_observations3 <- nrow(covid)
num_variables3 <- ncol(covid)

# Create a table with basic information
data_info3 <- data.frame(
  Metric = c("Number of Columns (Variables)", "Number of Rows (Observations)"),
  Value = c(num_variables3, num_observations3)
)

# Display the table
print(data_info3)
```

### Key variables :

### 1. Total Death: Total number of deaths recorded in a county due to COVID-19. This is a key indicator of the pandemic’s severity in the region.

### 2. Total Confirmed: Total number of confirmed COVID-19 cases in the region. This variable shows the extent of the outbreak in the area.

### 3. Admin 2 FIPS Code: Federal Information Processing Standards (FIPS) code, a numeric identifier for counties. It helps in mapping and analyzing data geographically.





>>>>>>> 9e5743d4b17787f36c3c05ef4511b923b9c423d2
